Using numerical modelling to investigate temperature and wave exposure as drivers

### 2019-08-28 morphology_paper.Rmd
references finished. I will review shortly but otherwise discussion is finished.

---
title: Using numerical modelling to investigate temperature and wave exposure as drivers of macroalgal characteristics
author: "R. Coppin, C. Rautenbach, T. Ponton and A.J Smit"
geometry: margin=1in
output:
  word_document:
    reference_docx: frontiers_template.docx
papersize: A4
tables: yes
fontsize: 10pt
bibliography: refs_morphology_paper.bib
---

@@

Temperature and wave exposure have been shown to play a role in driving kelp morphology. For example, a study by @Serisawa2002-dc compared the morphology of Ecklonia cava Kjellman growing in warm temperate and cool temperate morphologies. The results showed that wrinkles in the blade seem to be a characteristic of warm temperate regions. However this study did not take into account the interaction with other environmental variables such as wave exposure. Wave exposure has also been shown to be the main drivers of morphological adaptation across various kelp species. Past research has shown that in highly wave exposed areas kelp morphology tends to take on characteristics which reduce drag, increase strength and increase flexibility [@Denny2002-ob]. For example, a study by @Wernberg2005c examined the consistency of wave exposure as a driver of *Ecklonia radiata* (C. Agardh) J. Agardh across a broad geographic range, and  showed trends towards drag-reducing (small size, narrow laterals and blades, low spinosity) and increased strength (large holdfast, thick stipe and thick blades and lamina) at high wave exposure sites. The morphological adaptation to wave exposure must be kept in balance with other important processes, such as nutrient assimilation and light absorption, which are dependent on the amount of surface area of the blades. Therefore, there is a trade-off between reducing overall drag and maintaining nutrient and photosynthetic ability.  The morphological adapatibility of kelps is driven by a combination of environmental factors that in turn do not act independently of one another. A study by @Wernberg2003-mp investigated the morphology of *E. radiata* in order to quantify the morphological variation and whether it was dependent on spatial differences along the Australasian coast. They found no correlation between spatial distance and morphological similarity and rather the morphology of kelps was representative of multiple environmental forcings on different morphological characters at different spatial scales [@Wernberg2003-mp]. Due to the complex effects of environmental drivers on kelp morphology, one can also expect differences in morphology between deep and shallow water populations of the same species. Wave exposure may be greater in shallower environments, or it may be reduced through damping effects of kelp further offshore and natural barriers such as rock outcrops; also, seawater temperature may be higher in shallow water environments due to higher solar irradiance.

Other species of kelp, such as *Laminaria pallida* and *Ecklonia maxima* (Osbeck) Papenfuss which are important habitat forming seaweeds that exist around the coast of South Africa, offer a unique opportunity to investigate the drivers of macroalgal morphological characteristics between deep and shallow water environments. Although both these species exist in the subtidal, *L. pallida* dominates deeper waters while *E. maxima* forms dense surface canopies in from the subtidal to shallower waters. Therefore, *E. maxima* is exposed to variations in wave, swell and temperature, while *L. pallida* is exposed to variations in waves and temperature. @Molloy1996-la investigated the effect of depth and wave exposure on *L. pallida* at different depths and wave exposure and showed that depth had a greater effect than wave exposure when considering all the morphological characteristics; when considering individual characteristics, however, wave exposure had the most significant effect on blade thickness. Another study by @Rothman2017-pn investigated the changes in morphology in shallow populations of *L. pallida* and *E. maxima* along the South African coastline and into Namibia; *E. maxima* exhibited no morphological changes along the coast but the stipes of *L. pallida* become increasingly hollow further north along the coastline. They suggested that turbidity was the environmental driver responsible for this change.

Measures of wave exposure in ecological studies often incorporate integrative measures of hydrodynamic conditions at a particular site which are based on cartographical models of wave exposure. Cartographical models of wave exposure are regarded as 'fetch-based-models' which measure the length of open water associated with a particular site in a straight line, and are regarded as simple measures of wave exposure. Advances in numerical modelling based on physical/linear wave theory incorporate more complexity (wind forcing, wave-wave interactions, wave breaking, diffraction and variation in wave direction) into the models and allow for a quantitave, reproducible approach for measuring wave exposure. Currently, no ecological studies investigating macroalgal morphological characteristics in relation to the wave environment have incorporated 3D spectral numerical modelling. Previous research investigating drivers of kelp morphological characteristics used qualitative estimates of wave exposure and did not consider other morphological characteristics. Furthermore, possible differences in morphology between shallow and deep water populations have not been considered in previous research which ignores the effects of wave damping. The aim of this study is, therefore, to understand how temperature and wave exposure can influence the morphology in two species of kelps around South Africa, as well as between deep and shallow populations of the same species. This will be achieved by initially understanding the variation in temperature and waves using advanced numerical models. These numerical models will then be used as a basis for investigating the consequences for morphological chracteristics of *E. maxima* and *L. pallida* using statistical tools.

@@

Sites were chosen to represent an array of environmental gradients, as indicated in Fig. 1. St. Helena Bay and Betty's Bay constituted the north western and south eastern boundary sites, respectively. These sites are roughly 300 km apart, and lie within separate marine provinces and cover a cline in wave exposure and temperature. Study sites span across the majority of the south-west coast, in varying thermal and wave energy regimes. The region is dominated by kelp communities that persist in contrasting abiotic environments. The west coast region has been termed as a cool temperate region, which is defined as a region where mean monthly temperatures are always above 10°C and always below 15°C [@Smit2013a]. East of Cape Point marks the beginning of an overlap or transition area, which is also referred to as the Benguela-Agulhas Transition Zone [@Smit2013a]. The Agulhas Marine Province is characterised by a wide temperature range of up to 7°C difference between mean monthly temperatures between summer and winter and is classified as warm temperate region [@Smit2013a]. Due to the Cape Peninsula temperate latitude, winter months bring an increased frequency of frontal depressions that originate from the Southern Ocean [@Reason2006]. These low pressures are joined by large swells with increased wave energy. The nearshore environment, with the accompanied biota, therefore experiences high wave energy events, with increased frequency in winter [@Veitch2019]. The large peninsula acts as an obstruction for large south westerly swells, providing decreased wave energy along the west side of False Bay [@Shipley1964]. Conversely, the west coast of Cape Point is battered by these large swells. Multiple sites, therefore, exist where kelps grow in diverse temperature and wave energy climates, in close proximity.

@@

The topography and elevation along the Cape Peninsula channel and shield winds along False Bay. This is however absent in winter, where strong northerly winds are prevalent from St. Helena Bay to Betty's Bay [@Field1980a; @Jury1985-nr; @Andrews1980; @Jury1980-ov]. Patterns are also evident in the wave data. Seasonal variations in significant wave height (Hs), peak period and wave direction are also present. The direction of the dominant swell changes to the south west in winter, generated by strong low pressures that originate from the southern ocean [@Reason2006], which False Bay is sheltered from [@Shipley1964; @Atkins1970-bx; @Dufois2012-vy]. In summer, these swells rotate anticlockwise and are able to enter False Bay, providing an increased variability of Hs and Tp in this region. It should be noted that what is classified as sheltered around the South African coastline (a high energy coastline) might be classified as exposed in other regions of the world [@Norderhaug2012-ks; @Leliaert2000a]. It is due to the near consistent south-westerly swell and the complex orography around the peninsula that the wave energy distribution around the Cape Peninsula varies significantly over a small geographical area.

@@

```{r rda eck plots, echo=FALSE, fig.height=7, fig.width=8, message=FALSE, warning=FALSE}
ggarrange(lam_rda, rda_eck_deep, rda_eck_shallow, common.legend = TRUE, legend = "bottom", labels = "AUTO",ncol = 2, nrow = 2)
```

@@

This study considered important drivers of kelp morphology that have been identified in previous research [@FowlerWalker2006; @Miller2011-pm; @Fowler-Walker2005; @Stewart2009; @Wing2007-tv; @Molloy1996-la; @Pedersen2012-vr; @Wernberg2005c], namely temperature and wave exposure, around the Cape Peninsula on two prolific kelp species. Unlike previous research, this study used a numerical model to provide the various variables of wave exposure. We also considered wind in addition to temperature and wave variables, since wind is an important component of wind-driven surface waves. The results show that specific variables of wave exposure are the main drivers of kelp morphology with temperature only playing a minor role. Furthermore, the investigation of differences in morphological characteristics between shallow and deep populations of *E. maxima * suggests that in low wave exposure environments the role of temperature as a morphological driver increases, such as in False Bay.

The complex geomorphology of the western Cape coastline creates an ideal natural laboratory for studies of kelps in interaction with their environment. The annual and seasonal patterns in both wind influence the thermal regime around the coastline by inducing upwelling. Upwelling brings deep, cool nutrient rich water to coastal areas and therefore induces a decrease in temperature [@Field1980a; @Andrews1980; @Gill1974-nt; @Cram1970-jw; @Rouault2010-gi; @Blanke2002]. The wind speed and wind direction help drive upwelling in summer on the western side of the Cape peninsula, where southerly winds blow parallel to the coast and trigger upwelling [@Field1980a; @Rouault2010-gi]. This is not the case for False Bay which is shielded from the dominating wind and swell the Cape Peninsula is subject to, with the exception being the winter months where upwelling becomes more favourable for the region due to the change in swell and wind direction, causing lower temperatures to dominate during this time [@Rouault2010-gi; @Andrews1980]. Increased wind speed at sites along the west side of the peninsula in a southerly direction trigger upwelling events along the western side of the peninsula [@Rouault2010-gi; @Field1980a]. We however see a possible lack of upwelling events for sites such as Betty’s Bay and Batsata Rock.

We provide strong support that variations in the environmental variables, particularly wave exposure variables, are driving kelp morphological characteristics around the Cape Peninsula. Morphological adaptation due to water motion may manifest in a number of ways in high wave energy environments. For instance, reduction of blade thickness, blade elongation, increase of stipe length, increase in stipe circumference and force of attachment ([@FowlerWalker2006; @Wernberg2005c; @Friedland1995-vi; @Denny2002; @Bekkby2014; @Denny1997; Denny2002-ob]) have been suggested in previous studies. Although this study did not measure force of attachment, other morphological responses to temperature and wave variables were evident. Species-specific responses are evident to both wave exposure and temperature. In cool-temperate environments *L. pallida* tended to show increases in certain morphological characteristics (stipe mass, stipe length and stipe diameter) while in the warm-temperate environments these were significantly lower. This was also true for deep *E. maxima* populations, which were significantly taller and thinner for kelp in the warm-temperate compared to kelp in the cool-temperate. This has been documented previously by ?????, and show that the 'slender' kelp morphology is a response to warmer temperatures. In warmer environments kelp metabolism increases, using more nutrients than under cooler conditions causing the kelp to become slender **more needed here**. It should be noted however, that from these analyses that in general the cool-temperate region is more exposed to waves than that of the warm-temperate region. Therefore, the significantly larger morphological characteristics for kelp from the cool-temperate region may overlap with responses to wave exposure which was exhibited strongly in both species. When grouped by wave exposure category, *L. pallida* characteristics showed a significant increase in length for the 'Extremely sheltered' category (6-10 kW/m; figure ??) which maybe a 'go-with-the-flow' tactic. Kelp are able to increase flexibility by increasing stipe length which increases the extension capabilities of kelp to a passing wave. However, increasing stipe length is only beneficial under lower exposure levels as a long stipe actually increases overall drag on the plant under higher exposure levels. This is reflected in Figure ?? which shows an overall reduction of *L pallida* morphological characteristics suggesting a size reducing tactic to cope with higher levels of wave exposure. The morphological characteristics of *E. maxima* show similar patterns when grouped to wave exposure. The results suggest that deep *E. maxima* populations exhibit a different response as well as a higher wave exposure threshold. Unlike *L. pallida*, *E. maxima* exhibits a size reducing tactic with in an increase in exposure level by significantly decreasing stipe mass, stipe length, stipe circumference, total length and frond length. This size reduction tactic has been shown before (citations ???) and is regarded as strategy to reduce overall drag on the plant. However, *E. maxima* morphological characteristics changed significantly as wave exposure increases. When grouped to the higher wave exposure levels, *E. maxima* morphology exhibited a strength and flexibility increasing trait (see figure ??). Furthermore, the morphological response of *E. maxima* to wave exposure only occurs at the 'Moderately exposed' (20-30 kW/m) level compared to *L. pallida* which exhibits a response at a lower wave exposure level. This suggests that *E. maxima* can tolerate higher exposure levels before having to exhibit a morphological response. The redundancy analysis performed confirms the patterns and responses observed as well as identifying specific temperature and wave variables as drivers of kelp morphology. Kelp morphology characteristics are largely wave driven for both species but differ in terms of specific temperature and wave metrics. For instance, both species morphological characteristics are determined by swell direction (dir_median), standard deviation in peak period (tp_sd) and mean significant swell height (hs_mean), but *L. pallida* temperature driver is mean temperature (mean_temp) and *E. maxima* by minimum temperature (min_temp). Minimum temperature as an important metric for *E. maxima* may be related to nutrient availability. Low temperatures are often associated with upwelling events which bring cool, nutrient rich water into the nearshore. False Bay has comparativley little upwelling events compared to the western side of the peninsula, and so nutrients may be a limiting factor for *E. maxima* populations within False Bay (see figure ??). The kelp *L. pallida* is a sub-canopy species where more efficent mixing of water within the water column occurs (ref ??). The difference in canopy type between the species may also be the reason why *E. maxima* is driven by multiple wave metrics compared to *L. pallida*. Since *E. maxima* is a canopy-kelp, it is will be exposed to all components of a wave compared to *L. pallida* which occurs deeper in the water column.

Annual maximum and winter mean temperatures influence stipe length and frond length in shallow *E. maxima* individuals located at sites within False Bay. This was not the case for deeper *E. maxima* individuals as well as *L. pallida* which were largely driven by wave variables. The shallow *E. maxima* individuals may not experience the same amount of wave energy as do deeper populations. This may be due to wave dampening from deeper populations of *E. maxima*, which attenuate the wave energy entering coastal environments [@Kobayashi1993-os; @Jackson1984; @Mork1996-md]. The morphological variations in characteristics for deep *E. maxima* populations are not seen for shallow *E. maxima* populations. The lack of variation in shallow *E. maxima* populations was also found in previous research by [@Rothman2017-pn] who showed that light was a significant driver. This influence of light may suggest photosynthetic ability and nutrient uptake as important drivers in shallow environments. Wave exposure may be reduced in shallow environments due to the wave damping effect of deep kelp and seaweed populations [@Kobayashi1993-os; @Mork1996-md; @Jackson1984], or rocky outcrops, thereby reducing the need to adapt morphological characteristics to reduce drag. Furthermore, the reduction in differences between deep and shallow *E. maxima *at wave exposed sites on the western of the peninsula compared to sites in False Bay provides further evidence that wave exposure is not an important driver of shallow *E. maxima *populations. Increased temperatures corresponded with a reduction in  stipe circumference, frond length, stipe length, total length and stipe mass resulting in slim or ‘skinny’ kelp individuals, which supports findings from previous research by [@Mabin2013-zs; @Serisawa2002-dc].

>>>>

---
title: Using numerical modelling to investigate temperature and wave exposure as drivers
  of macroalgal characteristics
author: "R. Coppin, C. Rautenbach, T. Ponton and A.J Smit"
fontsize: 10pt
output:
  word_document:
    reference_docx: frontiers_template.docx
  pdf_document: default
geometry: margin=1in
papersize: A4
tables: yes
bibliography: refs_morphology_paper.bib
---

@@

Temperature and wave exposure have been shown to play a role in driving kelp morphology. For example, a study by @Serisawa2002-dc compared the morphology of *Ecklonia cava* Kjellman growing in warm temperate and cool temperate morphologies. The results showed that wrinkles in the blade and reduced size to be a characteristic of warm temperate regions. However this study did not take into account the interaction with other environmental variables such as wave exposure. Wave exposure has also been shown to be the main drivers of morphological adaptation across various kelp species. Past research has shown that in highly wave exposed areas kelp morphology tends to take on characteristics which reduce drag, increase strength and increase flexibility [@Denny2002-ob; @Hurd2000; @Fowler-Walker2005, @Thomsen2004-yd]. For example, a study by @Wernberg2005c examined the consistency of wave exposure as a driver of *Ecklonia radiata* (C. Agardh) J. Agardh across a broad geographic range, and  showed trends towards drag-reducing (small size, narrow laterals and blades, low spinosity) and increased strength (large holdfast, thick stipe and thick blades and lamina) at high wave exposure sites. The morphological adaptation to wave exposure must be kept in balance with other important processes, such as nutrient assimilation and light absorption, which are dependent on the amount of surface area of the blades. Therefore, there is a trade-off between reducing overall drag and maintaining nutrient and photosynthetic ability. The morphological adapatibility of kelps is driven by a combination of environmental factors that in turn do not act independently of one another. A study by @Wernberg2003-mp investigated the morphology of *E. radiata* in order to quantify the morphological variation and whether it was dependent on spatial differences along the Australasian coast. They found no correlation between spatial distance and morphological similarity and rather the morphology of kelps was representative of multiple environmental forcings on different morphological characters at different spatial scales [@Wernberg2003-mp]. Due to the complex effects of environmental drivers on kelp morphology, one can also expect differences in morphology between deep and shallow water populations of the same species. Wave exposure may be greater in shallower environments, or it may be reduced through damping effects of kelp and other subtidal species [@Kobayashi1993-os; Mia2013; @Dubi1995]. Shallower environments are exposed to higher irradiance levels which increases temperatures [@Dellatorre2012; @Dunne2001] as well as an altered hydrodynamic environment which effect vertical mixing in the shallow nearshore.

The kelps *Laminaria pallida* and *Ecklonia maxima* (Osbeck) Papenfuss are important habitat forming seaweeds that exist around the coast of South Africa and offer a unique opportunity to investigate the drivers of macroalgal morphological characteristics between canopy and sub-canopy species, as well as between deep and shallow water environments. Although both these species exist in the subtidal, *L. pallida* dominates deeper waters while *E. maxima* forms dense surface canopies in from the subtidal to shallower waters. Therefore, *E. maxima* is exposed to variations in wind drive surface waves, swell and temperature, while *L. pallida* is exposed to variations in swell and temperature. @Molloy1996-la investigated the effect of depth and wave exposure on *L. pallida* at different depths and wave exposure and showed that depth had a greater effect than wave exposure when considering all the morphological characteristics (wave exposure alters with depth); when considering individual characteristics, however, wave exposure had the most significant effect on blade thickness. Another study by @Rothman2017-pn investigated the changes in morphology in shallow populations of *L. pallida* and *E. maxima* along the South African coastline and into Namibia; *E. maxima* exhibited no morphological changes along the coast but the stipes of *L. pallida* become increasingly hollow further north along the coastline. They suggested that turbidity was the environmental driver responsible for this change.

Measures of wave exposure in ecological studies often incorporate integrative measures of hydrodynamic conditions at a particular site which are based on cartographical models of wave exposure. Cartographical models of wave exposure are regarded as 'fetch-based-models' which measure the length of open water associated with a particular site in a straight line, and are regarded as simple measures of wave exposure. Advances in numerical modelling based on physical/linear wave theory incorporate more complexity (wind forcing, wave-wave interactions, wave breaking, diffraction and variation in wave direction) into the models and allow for a quantitave, reproducible approach for measuring wave exposure. Currently, no ecological studies investigating macroalgal morphological characteristics in relation to the wave environment have incorporated 3D spectral numerical modelling. Previous research investigating drivers of kelp morphological characteristics have often used qualitative estimates of wave exposure and did not consider other morphological characteristics. Furthermore, wind is an important component of surface gravity waves [@Holthuijsen2010-vl] and previous work has not differentiated between surface gravity waves and swell waves, which the current work will aim to achieve. Other considerations such as possible differences in morphology between sub-canopy and canopy species, shallow and deep water populations have not been considered in previous research which ignores the effects of wave damping. The aim of this study is, therefore, to understand how temperature and wave exposure can influence the morphology in two species of kelps around South Africa, as well as between deep and shallow populations of the same species. This will be achieved by initially understanding the variation in temperature and waves using an advanced numerical model. The numerical model will then be used as a basis for investigating the consequences for morphological chracteristics of *E. maxima* and *L. pallida* using multiple statistical tools.

@@

Sites were chosen to represent an array of environmental gradients, as indicated in Fig. 1. St. Helena Bay and Betty's Bay constituted the north western and south eastern boundary sites, respectively. These sites are roughly 300 km apart, and lie within separate marine provinces and cover a cline in wave exposure and temperature. Study sites span across the majority of the south-west coast, in varying thermal and wave energy regimes. The region is dominated by kelp communities that persist in contrasting abiotic environments. The west coast region has been termed as a cool temperate region, which is defined as a region where mean monthly temperatures are always above 10°C and always below 15°C [@Smit2013a]. East of Cape Point marks the beginning of an overlap or transition area, which is also referred to as the Benguela-Agulhas Transition Zone [@Smit2013a]. The Agulhas Marine Province is characterised by a wide temperature range of up to 7°C difference between mean monthly temperatures between summer and winter and is classified as warm temperate region [@Smit2013a].

@@

Due to the Cape Peninsula temperate latitude, winter months bring an increased frequency of frontal depressions that originate from the Southern Ocean [@Reason2006]. These low pressures are joined by large swells with increased wave energy. The nearshore environment, with the accompanied biota, therefore experiences high wave energy events, with increased frequency in winter [@Veitch2019]. The large peninsula acts as an obstruction for large south westerly swells, providing decreased wave energy along the west side of False Bay [@Shipley1964]. Conversely, the west coast of Cape Point is battered by these large swells. Multiple sites, therefore, exist where kelps grow in diverse temperature and wave energy climates, in close proximity. The topography and elevation along the Cape Peninsula channel and shield winds along False Bay. This is however absent in winter, where strong northerly winds are prevalent from St. Helena Bay to Betty's Bay [@Field1980a; @Jury1985-nr; @Andrews1980; @Jury1980-ov]. Patterns are also evident in the wave data. Seasonal variations in significant wave height (Hs), peak period and wave direction are also present. The direction of the dominant swell changes to the south west in winter, generated by strong low pressures that originate from the southern ocean [@Reason2006], which False Bay is sheltered from [@Shipley1964; @Atkins1970-bx; @Dufois2012-vy]. In summer, these swells rotate anticlockwise and are able to enter False Bay, providing an increased variability of Hs and Tp in this region. It should be noted that what is classified as sheltered around the South African coastline (a high energy coastline) might be classified as exposed in other regions of the world [@Norderhaug2012-ks; @Leliaert2000a]. It is due to the near consistent south-westerly swell and the complex orography around the peninsula that the wave energy distribution around the Cape Peninsula varies significantly over a small geographical area.

@@

```{r rda eck plots, echo=FALSE, fig.height=8, fig.width=8, message=FALSE, warning=FALSE}
ggarrange(lam_rda, rda_eck_deep, rda_eck_shallow, common.legend = TRUE, legend = "bottom", labels = "AUTO",ncol = 2, nrow = 2)
```

@@

This study considered important drivers of kelp morphology that have been identified in previous research [@FowlerWalker2006; @Miller2011-pm; @Fowler-Walker2005; @Stewart2009; @Wing2007-tv; @Molloy1996-la; @Pedersen2012-vr; @Wernberg2005c; @Thomsen2004-yd], namely temperature and wave exposure, around the Cape Peninsula on two prolific kelp species. The complex geomorphology of the western Cape coastline creates an ideal natural laboratory for studies of kelps in interaction with their environment. Unlike previous research, this study used a numerical model to provide the various variables of wave exposure. We also considered wind in addition to temperature and wave variables, since wind is an important component of wind-driven surface gravity waves [@Holthuijsen2010-vl]. The results show that specific variables of wave exposure are the main drivers of kelp morphology with temperature only playing a minor role. Furthermore, the investigation of differences in morphological characteristics between shallow and deep populations of *E. maxima * suggests that in low wave exposure environments the role of temperature as a morphological driver increases, such as in False Bay.

The annual and seasonal patterns in both wind influence the thermal regime around the coastline by inducing upwelling. Upwelling brings deep, cool nutrient rich water to coastal areas and therefore induces a decrease in temperature [@Field1980a; @Andrews1980; @Gill1974-nt; @Cram1970-jw; @Rouault2010-gi; @Blanke2002]. The wind speed and wind direction help drive upwelling in summer on the western side of the Cape peninsula, where southerly winds blow parallel to the coast and trigger upwelling [@Field1980a; @Rouault2010-gi]. This is not the case for False Bay which is shielded from the dominating wind and swell the Cape Peninsula is subject to, with the exception being the winter months where upwelling becomes more favourable for the region due to the change in swell and wind direction, causing lower temperatures to dominate during this time [@Rouault2010-gi; @Andrews1980]. Increased wind speed at sites along the west side of the peninsula in a southerly direction trigger upwelling events along the western side of the peninsula [@Rouault2010-gi; @Field1980a].

We provide strong support that variations in the environmental variables, particularly wave exposure variables, are driving kelp morphological characteristics around the Cape Peninsula. Morphological adaptation due to water motion may manifest in a number of ways in high wave energy environments. For instance, reduction of blade thickness, blade elongation, increase of stipe length, increase in stipe circumference and force of attachment [@FowlerWalker2006; @Wernberg2005c; @Friedland1995-vi; @Denny2002; @Bekkby2014; @Denny1997; @Denny2002-ob] have been identified in previous studies. Although this study did not measure force of attachment, other morphological responses to temperature and wave variables were evident. Species-specific responses are evident in both wave exposure and temperature. In cool-temperate environments *L. pallida* tended to show increases in certain morphological characteristics (stipe mass, stipe length and stipe diameter) while in the warm-temperate environments these were significantly lower. This was also true for deep *E. maxima* populations, which had longer, thinner stipes for kelp in the warm-temperate compared to kelp in the cool-temperate. Reduction in certian morphological charateristics has been attributed to temperature by @Serisawa2002-dc in the kelp *E. cava*, which was smaller and shorter in warmer sites compared to cooler sites. The reduction in size of adult may be a response to low nutrient conditions, which has been shown to reduce growth rate and overall morphology [@Simonson2015]. Warmer temperatures are associated with low nutrient concentrations [@Waldron1992] and the low frequency of upwelling conditions in False Bay (low nutrient supply) coupled with warmer temperatures may be a contributing factor. It should be noted however, that from these analyses that in general the cool-temperate region is more exposed to waves than that of the warm-temperate region. Therefore, the significantly larger morphological characteristics for kelp from the cool-temperate region may overlap with responses to wave exposure which was exhibited strongly in both species. When grouped by wave exposure category, *L. pallida* characteristics showed a significant increase in length for the 'Extremely sheltered' category (6-10 kW/m; figure ??) compared to the fully sheltered category, which maybe a 'go-with-the-flow' tactic. Kelp are able to increase flexibility by increasing stipe length which increases the extension capabilities of kelp to a passing wave [@Denny2002-ob; @Hurd2000]. However, increasing stipe length is only beneficial under lower exposure levels as a long stipe actually increases overall drag on the plant under higher exposure levels [@Denny2002-ob; @Denny1997]. This is reflected in Figure ?? which shows an overall reduction of *L pallida* morphological characteristics suggesting a size reducing tactic to cope with higher levels of wave exposure. The morphological characteristics of *E. maxima* show similar patterns when grouped to wave exposure. The results suggest that deep *E. maxima* populations exhibit a different response as well as a higher wave exposure threshold. Unlike *L. pallida*, *E. maxima* exhibits a size reducing tactic with in an increase in exposure level by significantly decreasing stipe mass, stipe length, stipe circumference, total length and frond length. A magnitude related reponse has been suggested by @Wernberg2010 who investigated the temporal and spatial variations of various environmental drivers of the kelp *E. radiata*. The authors identified wave exposure as the most important driver of kelp morphological characteristics and that the type of response elicited is dependent on the magnitude of wave exposure.

A size reduction tactic has been shown before [@FowlerWalker2006; @Denny2002-ob; @Hurd2000; @Fowler-Walker2005] and is regarded as strategy to reduce overall drag on the plant. However, *E. maxima* morphological characteristics changed significantly as wave exposure increases. When grouped to the higher wave exposure levels, *E. maxima* morphology exhibited a strength and flexibility increasing trait (see figure ??). Furthermore, the morphological response of *E. maxima* to wave exposure only occurs at the 'Moderately exposed' (20-30 kW/m) level compared to *L. pallida* which exhibits a response at a lower wave exposure level. This suggests that *E. maxima* can tolerate higher exposure levels before having to exhibit a morphological response. The redundancy analysis performed confirms the patterns and responses observed as well as identifying specific temperature and wave variables as drivers of kelp morphology. Kelp morphology characteristics are largely wave driven for both species but differ in terms of specific temperature and wave metrics. For instance, both species morphological characteristics are determined by swell direction (*dir_median*), standard deviation in peak period (*tp_sd*) and mean significant swell height (*hs_mean*), but *L. pallida* temperature driver is mean temperature (*mean_temp*) and *E. maxima* by minimum temperature (*min_temp*). Minimum temperature as an important metric for *E. maxima* may be related to nutrient availability. Low temperatures are often associated with upwelling events which bring cool, nutrient rich water into the nearshore. False Bay has comparativley little upwelling events compared to the western side of the peninsula, and so nutrients may be a limiting factor for *E. maxima* populations within False Bay (see figure ??) and hence the identification of minimum temperature (*min_temp*) as a driver. The difference in canopy type between the species may also be the reason why *E. maxima* is driven by multiple wave metrics compared to *L. pallida*. Since *E. maxima* is a canopy-kelp, it is  exposed to all components of a wave compared to *L. pallida* which occurs deeper in the water column. Mean temperature identified as a driver for *L. pallida* may be related to diurnal temperature fluctuations in the water column. Solar heating of the water surface in combination with wind-driven transport causes fluctuations within the water coloumn, which can occur daily or seasonally [@Kaplan2003].

Annual maximum and winter mean temperatures influence stipe length and frond length in shallow *E. maxima* individuals located at sites within False Bay. This was not the case for deeper *E. maxima* individuals as well as *L. pallida* which were largely driven by wave variables. The shallow *E. maxima* individuals may not experience the same amount of wave energy as do deeper populations. This may be due to wave dampening from deeper populations of *E. maxima*, which attenuate the wave energy entering coastal environments [@Kobayashi1993-os; @Jackson1984; @Mork1996-md]. The morphological variations in characteristics for deep *E. maxima* populations are not seen for shallow *E. maxima* populations. The lack of variation in shallow *E. maxima* populations was also found in previous research by [@Rothman2017-pn] who showed that light was a significant driver. This influence of light may suggest photosynthetic ability and nutrient uptake as important drivers in shallow environments. Wave exposure may be reduced in shallow environments due to the wave damping effect of deep kelp and seaweed populations [@Kobayashi1993-os; @Mork1996-md; @Jackson1984] thereby reducing the need to adapt morphological characteristics to reduce drag. Furthermore, the reduction in differences between deep and shallow *E. maxima * at wave exposed sites on the western of the peninsula compared to sites in False Bay provides further evidence that wave exposure is not an important driver of shallow *E. maxima* populations. Increased temperatures corresponded with a reduction in stipe circumference, frond length, stipe length, total length and stipe mass resulting in slim or ‘skinny’ kelp individuals, which supports findings from previous research by [@Mabin2013-zs; @Serisawa2002-dc]. These findings were confirmed by the redundancy analysis which showed that contribution of wave variables as a driver of kelp morphological characterstics was lower compared to temperature variables. The sites do cluster by region but almost overlap entirely, suggesting that wave variables are not the main driver. Although wave variables were not identified as an important driver is is important to note the wave variables identified in the RDA differ to deep *E. maxima*. The identification of median wind direction (*dirw_median*) suggests that surface gravity waves as a driver compared to swell waves (*dir_median*) identifiend for *L. pallida* and *E. maxima*. The identification of maximum temperature as a driver of kelp morphological characteristics may be due to depth and irradiance. Temperature conditions along the shallow depths of the coastline have been shown to be warmer than deeper waters due to a combination of depth and higher irradiance in shallow coastal environments as water is able to heat up more effectivley than deeper waters.
